---
title: Why Banning International Travel has Little Effect on Coronavirus Spread
author: Will Dearden
date: '2020-03-14'
slug: travel-coronavirus
categories: []
tags:
  - R
  - coronavirus
image:
  caption: ''
  focal_point: ''
---

In my [previous post](/flattening-the-curve/), I covered the SIR model and why "social distancing" is so effective at slowing disease spread and reducing healthcare system burden. In this post I'm going to cover another intervention, banning international travel.

To do this, I'm going to have to extend the SIR model. One of the reasons that the coronavirus spreads so easily is that people can be infectious for several days while still being asymptomatic. So, instead of assuming three states (susceptible, infectious, and recovered), I'm going to assume the disease can follow five states: Susceptible (S), Asymptomatic (A), Hospitalized (H), Isolated (I), and Recovered (R). The Susceptible become infected but Asymptomatic. Then the Asymptomatic eventually become ill; some percent become ill enough to be Hospitalized and the rest become Isolated. The Hospitalized eventually heal enough to be Isolated. And eventually the Isolated become Recovered. I assume that only the Asymptomatic can spread the virus because both the Hospitalized and Isolated are following effective quarantine measures.

I will get to international travel later, but for now this model will allow us to break up Infected group to more accurately model spread dynamics. This model looks like

\begin{align*} 
\frac{dS}{dt} &= -\beta A S \\
\frac{dA}{dt} &= \beta A S - \kappa A \\
\frac{dH}{dt} &= \kappa \eta A - \tau H \\
\frac{dI}{dt} &= \tau H + \kappa (1 - \eta) A - \phi I \\
\frac{dR}{dt} &= \phi I
\end{align*}

We can interpret $1 / \kappa$ as the average number of days that a person is asymptomatic, $1 / \tau$ as the average number of days hospitalized, $\eta$ as the percent of people who require hospitalization, and $1 / \phi$ as the average length of time in isolation.

Putting this model in R and solving, we get:

```{r}
library(deSolve)
library(data.table)
library(ggplot2)

sahir_ode <- function(times, init, parms) {
    with(as.list(c(parms, init)), {
        dS <- -beta * A * S
        dA <- beta * A * S - kappa * A
        dH <- kappa * eta * A - tau * H
        dI <- tau * H + kappa * (1 - eta) * A - phi * I
        dR <- phi * I
        list(c(dS, dA, dH, dI, dR))
    })
}

asymptomatic_time <- 5
hospitalized_time <- 5
hospitalized_percent <- 0.2
isolated_time <- 14

R0 <- 2
beta <- R0 / asymptomatic_time
kappa <- 1 / asymptomatic_time
eta <- hospitalized_percent
tau <- 1 / hospitalized_time
phi <- 1 / isolated_time

us_asymptomatic <- 20000
us_population <- 3e11
fraction_asymptomatic_us <- us_asymptomatic / us_population

parms <- c(beta = beta, kappa = kappa, eta = eta, tau = tau, phi = phi)
init <- c(S = 1 - fraction_asymptomatic_us, A = fraction_asymptomatic_us, H = 0, I = 0, R = 0)
times <- seq(0, 200, length.out = 2000)

sahir_out <- lsoda(init, times, sahir_ode, parms)

sahir_dt <- as.data.table(sahir_out)

group_names <- c('Susceptible', 'Asymptomatic', 'Hospitalized', 'Isolated', 'Recovered')
names(sahir_dt) <- c('time', group_names)

sahir_dt <- melt(
    sahir_dt, 
    id.vars = c('time'), 
    measure.vars = group_names
)

ggplot(sahir_dt, aes(x = time, y = value, color = variable)) +
    geom_line(lwd = 1.5) +
    theme_minimal() +
    theme(text = element_text(size = 15)) +
    labs(x = 'Days', y = 'Fraction in Group', color = 'Group')
```

```{r echo=FALSE}
percent_hospitalized <- sprintf("%1.1f%%", 100*max(sahir_dt[variable == "Hospitalized"]$value))
```

I should note that even though the peak percent hospitalized here looks small, it is `r percent_hospitalized`. There are less than a million hospital beds in the US, which is about 0.3% of the population. So even `r percent_hospitalized` requiring hospitalization would result in hospitals because vastly overrun.

### International Travel

Now I'm going to model international travel. Let's assume that there are two populations of roughly equal size. One has a head start on the other in terms of number of infected. Let's assume it has three times as many asymptomatic people as the US. Each population follows the same dynamics as before but I'll allow for travel between the two populations. Only the Susceptible, Asymptomatic, and Recovered travel. I'll set a parameter $\rho$, which is equal to the percent of people traveling each day. I will try out $\rho = 0.002$ (equivalent to 0.2% of the population taking international travel daily) and $\rho = 0$. 0.2% is roughly the percent of the US taking international travel daily so is a reasonable parameter for this model.

The more complicated model is then

\begin{align*} 
\frac{dS_1}{dt} &= -\beta A_1 S_1 + \rho (S_2 - S_1) \\
\frac{dA_1}{dt} &= \beta A_1 S_1 - \kappa A_1 + \rho (A_2 - A_1) \\
\frac{dH_1}{dt} &= \kappa \eta A_1 - \tau H_1 \\
\frac{dI_1}{dt} &= \tau H_1 + \kappa (1 - \eta) A_1 - \phi I_1 \\
\frac{dR_1}{dt} &= \phi I_1  + \rho (R_2 - R_1)\\
\end{align*}

\begin{align*}
\frac{dS_2}{dt} &= -\beta A_2 S_2  + \rho (S_1 - S_2)\\
\frac{dA_2}{dt} &= \beta A_2 S_2 - \kappa A_2  + \rho (A_1 - A_2)\\
\frac{dH_2}{dt} &= \kappa \eta A_2 - \tau H_2 \\
\frac{dI_2}{dt} &= \tau H_2 + \kappa (1 - \eta) A_2 - \phi I_2 \\
\frac{dR_2}{dt} &= \phi I_2 + \rho (R_1 - R_2)
\end{align*}

```{r}

sahir_travel_ode <- function(times, init, parms) {
    with(as.list(c(parms, init)), {
        dS1 <- -beta * A1 * S1 + rho * (S2 - S1)
        dA1 <- beta * A1 * S1 - kappa * A1 + rho * (A2 - A1)
        dH1 <- kappa * eta* A1 - tau * H1
        dI1 <- tau * H1 + kappa * (1 - eta) * A1 - phi * I1
        dR1 <- phi * I1 + rho * (R2 - R1)
        
        dS2 <- -beta * A2 * S2 + rho * (S1 - S2)
        dA2 <- beta * A2 * S2 - kappa * A2 + rho * (A1 - A2)
        dH2 <- kappa * eta * A2 - tau * H2
        dI2 <- tau * H2 + kappa * (1 - eta) * A2 - phi * I2
        dR2 <- phi * I2 + rho * (R1 - R2)
        
        list(c(dS1, dA1, dH1, dI1, dR1, dS2, dA2, dH2, dI2, dR2))
    })
}

simulate_sahir_travel <- function(rho) {
    fraction_asymptomatic_europe <- fraction_asymptomatic_us * 3
    
    parms <- c(beta = beta, kappa = kappa, tau = tau, phi = phi, rho = rho)
    init <- c(
        S1 = 1 - fraction_asymptomatic_us, A1 = fraction_asymptomatic_us, H1 = 0, I1 = 0, R1 = 0,
        S2 = 1 - fraction_asymptomatic_europe, A2 = fraction_asymptomatic_europe, H2 = 0, I2 = 0, R2 = 0
    )
    times <- seq(0, 200, length.out = 2000)
    
    sahir_out <- lsoda(init, times, sahir_travel_ode, parms)
    
    sahir_dt <- as.data.table(sahir_out)

    indexed_names <- as.vector(outer(group_names, c(': US', ': Europe'), FUN = paste0))
    names(sahir_dt) <- c('time', indexed_names)
    
    sahir_dt <- melt(
        sahir_dt, 
        id.vars = c('time'), 
        measure.vars = indexed_names
    )
    
    sahir_dt
}

sahir_no_travel <- simulate_sahir_travel(0)
sahir_travel <- simulate_sahir_travel(0.002)

sahir_no_travel_us <- sahir_no_travel[grepl('US$', sahir_no_travel$variable)]
sahir_travel_us <- sahir_travel[grepl('US$', sahir_travel$variable)]
sahir_dt <- rbindlist(list("No travel" = sahir_no_travel_us, "Travel" = sahir_travel_us), idcol = "travel")

sahir_dt$variable <- sub(": US$", '', sahir_dt$variable)
sahir_dt$variable <- factor(sahir_dt$variable, levels = group_names)

sahir_dt$travel <- factor(sahir_dt$travel, levels = c('Travel', 'No travel'))

ggplot(sahir_dt, aes(x = time, y = value, color = variable, linetype = travel)) +
    geom_line(lwd = 0.5) + 
    theme_minimal() +
    theme(text = element_text(size = 15)) +
    labs(
        x = 'Days', 
        y = 'Fraction in Group', 
        color = 'Group', 
        linetype = 'Travel',
        title = "Effect of international travel on COVID-19 spread"
    )
```

As we can see, there is minimal effect of banning international travel on reducing the spread of COVID-19. The reason is that it has already spread to the United States. Even though other countries have a head start, allowing travel only shifts the trajectory of the disease spread up a few days.

There is one main drawback to this model: the effect of travel itself. When people fly, they are traveling in close quarters with others, in recycled air, and with reduced immune systems. So the act of travel itself temporarily spikes $\beta$ for those traveling. Nevertheless, this model shows that the most effective way to reduce the spread of the coronavirus is consistent and extended social distancing.